---
title: "Figure 1&S1"
output: html_document
date: "2026-05-14"
---

```{r 1. Load All Required Packages}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 6)

library(tidyverse)
library(readxl)
library(openxlsx)
library(fs)
library(matrixStats)
library(ggrepel)
library(limma)
library(plotly)
library(pheatmap)
library(VIM)
library(sva)
library(RColorBrewer)
library(ggforce)
library(patchwork)
library(corrplot)
library(scales)
library(knitr)
library(reshape2)
library(ggplot2)
library(dplyr)
library(readxl)
library(openxlsx)
library(ggplot2)
library(ggrepel)
library(patchwork)

source(file.path(knitr::current_input(dir = TRUE), "..", "_common", "repro_paths.R"))
paths <- repro_paths("Code_chapter01")
repo_root <- paths$repo_root
input_dir <- paths$input_dir
data_dir <- paths$results_dir
fig_dir <- paths$results_dir
folder_path <- file.path(input_dir, "Data processing and Quality assessment")

# --- Parameter Settings ---
file_pattern <- "\\.xlsx$"     # File matching pattern
protein_id_col <- "Accession"  # Protein ID column name
missing_threshold <- 30        # Missing value filtering threshold (%)

# Normalization parameters
dmso_pattern <- "DMSO"
pool_pattern <- "pool"

# Missing value imputation parameters
impute_width <- 0.3 
impute_shift <- 1.8
```

#  STEP 1: Data Import and Merging

```{r}
# Get file list
file_list <- dir_ls(folder_path, regexp = file_pattern)

if (length(file_list) == 0) stop("No Excel files found. Please check folder_path.")
cat("Found", length(file_list), "files.\n")

# Read and merge
all_data_list <- list()

for (file in file_list) {
  batch_name <- tools::file_path_sans_ext(basename(file))
  temp_data <- read_excel(file)
  
  # Rename columns: Add BatchID prefix
  new_names <- sapply(names(temp_data), function(col) {
    if (col == protein_id_col) return(protein_id_col)
    else return(paste(batch_name, col, sep = "_"))
  })
  names(temp_data) <- new_names
  all_data_list[[batch_name]] <- temp_data
}

# Merge all data
protein_data <- purrr::reduce(all_data_list, full_join, by = protein_id_col)
cat("Merged data dimensions:", paste(dim(protein_data), collapse = " x "), "\n")

# Generate Sample Info
sample_info <- tibble(FullSampleName = names(protein_data)[-1]) %>%
  mutate(
    Batch = str_extract(FullSampleName, "^[^_]+"),
    SampleType = str_extract(FullSampleName, "[^_]+$")
  )

# Save intermediate results
write.xlsx(protein_data, file.path(data_dir, "01_Raw_Data_Merged_10219_841.xlsx"))
write.xlsx(sample_info, file.path(data_dir, "02_Sample_Info.xlsx"))
```

# STEP 2: Contaminant Filtering

```{r filter_contaminants}
# Contaminant list file path
contaminant_file_path <- file.path(input_dir, "Dec2022_Sample specific FASTA protein list.xlsx")

initial_count <- nrow(protein_data)
protein_data <- protein_data %>% 
  filter(!str_starts(!!sym(protein_id_col), "Cont"))

cat(paste0("Remaining proteins after removing 'Cont' prefix: ", nrow(protein_data), 
           " (Removed ", initial_count - nrow(protein_data), ")\n"))

contaminant_file <- file.path(input_dir, "Dec2022_Sample specific FASTA protein list.xlsx")

if (file.exists(contaminant_file)) {
  
  cat(paste("\nReading contaminant list:", contaminant_file, "...\n"))
  
  # Read Excel
  contaminants_library <- read_excel(contaminant_file, sheet = "common 381")
  
  # Extract IDs and filter
  contaminant_ids <- contaminants_library$`Uniprot ID`
  
  # Calculate number to be removed
  n_to_remove <- sum(protein_data$Accession %in% contaminant_ids)
  
  # Perform filtering
  protein_data <- protein_data[!protein_data$Accession %in% contaminant_ids, ]
  
  cat(paste0("Removed ", n_to_remove, " contaminant proteins based on list.\n"))
  
  # Clean up temporary variables
  rm(contaminants_library, contaminant_ids)
  
} else {
  warning(paste("Contaminant list file not found:", contaminant_file, ". Only prefix filtering performed."))
}

# Output results
final_protein_count <- nrow(protein_data)
cat(paste0("\n>>> Final protein count after filtering: ", final_protein_count, "\n"))

# Save file to data_dir folder
save_file_name <- paste0("03_Contaminants_Removed_", final_protein_count, ".xlsx") 
write.xlsx(protein_data, file.path(data_dir, save_file_name))

# Environment cleanup
suppressWarnings(rm(protein_data_filtered, clean_expression_matrix, temp_data))
```

# STEP 3: Data Completeness Statistics

```{r data_completeness}
quant_mat <- as.matrix(protein_data[, -1])
n_compounds <- ncol(quant_mat)
n_nonmiss_per_protein <- rowSums(!is.na(quant_mat))

# 1. Protein count at different quantification fractions
fractions <- seq(0.1, 1, by = 0.1)
thresholds <- ceiling(fractions * n_compounds)
proteins_at_least <- sapply(thresholds, function(th) sum(n_nonmiss_per_protein >= th))

df_completeness <- data.frame(fraction = fractions, n_proteins = proteins_at_least)

p_completeness <- ggplot(df_completeness, aes(x = fraction, y = n_proteins)) +
  geom_col(fill = "#C7E0FB") +
  scale_x_continuous(labels = scales::percent) +
  labs(title = "Dataset Completeness", x = "Min % of Compounds Quantified", y = "Number of Proteins") +
  theme_bw() + theme(aspect.ratio = 1)
print(p_completeness)

# 2. Proteome depth per compound
n_proteins_per_compound <- colSums(!is.na(quant_mat))
df_depth <- data.frame(compound = colnames(quant_mat), n_proteins = n_proteins_per_compound)
stats_vec <- c(
  min    = min(df_depth$n_proteins, na.rm = TRUE),
  median = median(df_depth$n_proteins, na.rm = TRUE),
  mean   = round(mean(df_depth$n_proteins, na.rm = TRUE)),
  max    = max(df_depth$n_proteins, na.rm = TRUE)
)

stats_df <- data.frame(
  stat  = c("Min", "Median", "Mean", "Max"),
  value = c(stats_vec["min"], stats_vec["median"], stats_vec["mean"], stats_vec["max"])
)

subtitle_txt <- sprintf(
  "Min = %d, Median = %d, Mean = %d, Max = %d",
  round(stats_vec["min"]),
  round(stats_vec["median"]),
  stats_vec["mean"],
  round(stats_vec["max"])
)

cat("Proteome Depth Statistics per Compound:\n")
print(stats_vec)

p_depth <- ggplot(df_depth, aes(x = n_proteins)) +
  geom_histogram(binwidth = 100, fill = "#C7E0FB", color = "white") +
  geom_vline(
    data = stats_df,
    aes(xintercept = value, color = stat),
    linetype = "dashed",
    linewidth = 0.8,
    show.legend = TRUE
  ) +
  scale_color_manual(
    name   = "Statistic",
    values = c("Min" = "gray40", "Median" = "orange", "Mean" = "blue", "Max" = "red")
  ) +
  labs(
    title    = "Proteome Depth per Compound",
    subtitle = subtitle_txt,
    x = "Proteins Quantified",
    y = "Number of Compounds"
  ) +
  theme_bw() + theme(aspect.ratio = 1)

print(p_depth)
#write.xlsx(df_completeness, file.path(data_dir, "Plot01_df_completeness.xlsx"))
#write.xlsx(df_depth, file.path(data_dir, "Plot02_df_depth.xlsx"))
#ggsave(file.path(fig_dir, "01_p_completeness.pdf"), p_completeness, width = 7, height = 7)
#ggsave(file.path(fig_dir, "02_p_depth.pdf"), p_depth, width = 7, height = 7)
```

#  STEP 4: Missing Value Filtering

```{r filter_missing}
missing_rates <- rowSums(is.na(protein_data[, -1])) / ncol(protein_data[, -1]) * 100
remaining_proteins <- protein_data[missing_rates <= missing_threshold, ]

cat("Filtering proteins with missing rate >", missing_threshold, "%.\n")
cat("Remaining protein count:", nrow(remaining_proteins), "\n")

write.xlsx(remaining_proteins, file.path(data_dir, "04_Filtered_Missing_70percent_6682.xlsx"))
protein_data <- remaining_proteins # Update main data
```

# STEP 5 & 6: Within-Batch Normalization (TIC)

```{r tic_normalization}
long_data <- protein_data %>%
  pivot_longer(cols = -all_of(protein_id_col), names_to = "SampleFullName", values_to = "Intensity") %>%
  separate(SampleFullName, into = c("BatchID", "Channel"), sep = "_", extra = "merge")

# Calculate TIC
total_intensities <- long_data %>% group_by(BatchID, Channel) %>% 
  summarise(Total_Intensity = sum(Intensity, na.rm = TRUE), .groups = 'drop')
target_intensities <- total_intensities %>% group_by(BatchID) %>% 
  summarise(Target_Intensity = mean(Total_Intensity, na.rm = TRUE), .groups = 'drop')

# Apply normalization
long_data_normalized <- long_data %>%
  left_join(total_intensities, by = c("BatchID", "Channel")) %>%
  left_join(target_intensities, by = "BatchID") %>%
  mutate(Intensity_Normalized = Intensity * (Target_Intensity / Total_Intensity)) %>%
  dplyr::select(!!sym(protein_id_col), BatchID, Channel, Intensity = Intensity_Normalized)

cat("TIC Normalization completed.\n")

# Visualization (Sampling 10%)
set.seed(123)
proteins_sample <- sample(unique(long_data$Accession), size = floor(0.1 * nrow(protein_data)))
plot_data_tic <- long_data_normalized %>%
  filter(Accession %in% proteins_sample) %>%
  mutate(Log2_Intensity = log2(Intensity + 1))

p_tic <- ggplot(plot_data_tic, aes(x = BatchID, y = Log2_Intensity)) +
  geom_boxplot(outlier.shape = NA, fill = "lightblue") +
  labs(title = "Intensity Distribution After TIC Normalization") +
  theme_bw() + theme(axis.text.x = element_text(angle = 90))
print(p_tic)
```

# STEP 7 & 8: Between-Batch Normalization (DMSO Bridging)

```{r dmso_bridging}
log_intensity_long <- long_data_normalized %>%
  unite("SampleID", BatchID, Channel, sep = "_", remove = FALSE) %>%
  mutate(Log2_Intensity = log2(Intensity + 1))

# Convert back to wide matrix
log_intensity_matrix <- log_intensity_long %>%
  pivot_wider(id_cols = !!sym(protein_id_col), names_from = SampleID, values_from = Log2_Intensity) %>%
  column_to_rownames(var = protein_id_col)

# Calculate global DMSO mean
dmso_cols <- colnames(log_intensity_matrix)[str_detect(colnames(log_intensity_matrix), dmso_pattern)]
global_mean_dmso <- rowMeans(log_intensity_matrix[, dmso_cols], na.rm = TRUE)

# Perform correction
dmso_corrected_matrix <- log_intensity_matrix
all_batches <- unique(long_data_normalized$BatchID)

for (batch in all_batches) {
  cols_in_batch <- colnames(log_intensity_matrix)[str_starts(colnames(log_intensity_matrix), paste0(batch, "_"))]
  dmso_col_in_batch <- cols_in_batch[str_detect(cols_in_batch, dmso_pattern)]
  
  if (length(dmso_col_in_batch) == 1) {
    correction_vector <- global_mean_dmso - log_intensity_matrix[, dmso_col_in_batch]
    correction_vector[is.na(correction_vector)] <- 0
    # Correct all columns in this batch
    dmso_corrected_matrix[, cols_in_batch] <- log_intensity_matrix[, cols_in_batch] + correction_vector
  }
}

# write.xlsx(dmso_corrected_matrix, file.path(data_dir, "05_DMSO_Corrected_Matrix_Before_Imputation.xlsx"), rowNames = TRUE)

# Effect evaluation boxplot
plot_data_dmso <- as.data.frame(dmso_corrected_matrix) %>%
  pivot_longer(cols = everything(), names_to = "SampleID", values_to = "Log2_Intensity") %>%
  mutate(Batch = str_extract(SampleID, "^[^_]+")) %>%
  mutate(Batch_Num = as.numeric(str_extract(Batch, "\\d+"))) %>%
  mutate(Batch = fct_reorder(Batch, Batch_Num))

p_dmso_box <- ggplot(plot_data_dmso, aes(x = Batch, y = Log2_Intensity)) +
  geom_boxplot(outlier.shape = NA, fill = "#9e539d", alpha=0.6) +
  labs(title = "Intensity After DMSO Bridging Correction", x = "Batch") +
  theme_bw() + theme(axis.text.x = element_text(angle = 90))
print(p_dmso_box)
```

#  STEP 9: Combined Visualization of Within-Batch & DMSO Correction

```{r}
before_long <- log_intensity_long %>%
  dplyr::select(
    dplyr::all_of(protein_id_col),
    BatchID,
    Channel,
    Log2_Intensity
  ) %>%
  dplyr::mutate(Status = "Before Normalization")

after_long <- as.data.frame(dmso_corrected_matrix) %>%
  tibble::rownames_to_column(var = protein_id_col) %>%
  tidyr::pivot_longer(
    cols      = -dplyr::all_of(protein_id_col),
    names_to  = "SampleID",
    values_to = "Log2_Intensity"
  ) %>%
  tidyr::separate(
    col  = SampleID,
    into = c("BatchID", "Channel"),
    sep  = "_",
    extra = "merge"
  ) %>%
  dplyr::select(
    dplyr::all_of(protein_id_col),
    BatchID,
    Channel,
    Log2_Intensity
  ) %>%
  dplyr::mutate(Status = "After Normalization")

# Merge data
plot_data <- bind_rows(before_long, after_long) %>%
  mutate(
    Batch_Number = as.numeric(str_extract(BatchID, "\\d+")),
    BatchID_Sorted = fct_reorder(BatchID, Batch_Number),
    Status = factor(Status, levels = c("Before Normalization", "After Normalization")) 
  )

dmso_agg_boxplot_final <- ggplot(plot_data, aes(x = BatchID_Sorted, y = Log2_Intensity, fill = Status)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA) +
  coord_cartesian(ylim = quantile(plot_data$Log2_Intensity, c(0.05, 0.95), na.rm = TRUE)) +
  labs(
    title = "Evaluation of Batch Correction Effect using DMSO Bridging",
    subtitle = "Distribution of protein intensities across batches before and after normalization",
    x = "Batch ID",
    y = "Log2(Intensity)",
    fill = "Normalization Status" 
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "top"
  )

print(dmso_agg_boxplot_final)
```

# STEP 10: Missing Value Imputation (MinDet)

```{r imputation}
set.seed(123)
imputed_matrix <- as.matrix(dmso_corrected_matrix)

for (i in 1:ncol(imputed_matrix)) {
  vec <- imputed_matrix[, i]
  if (any(is.na(vec))) {
    stats <- c(mean = mean(vec, na.rm=T), sd = sd(vec, na.rm=T))
    if (!is.na(stats['sd']) && stats['sd'] > 0) {
      n_miss <- sum(is.na(vec))
      imp_mean <- stats['mean'] - (impute_shift * stats['sd'])
      imp_sd <- impute_width * stats['sd']
      imputed_matrix[is.na(vec), i] <- rnorm(n_miss, mean = imp_mean, sd = imp_sd)
    }
  }
}

dmso_corrected_matrix <- imputed_matrix # Update main matrix
# write.xlsx(dmso_corrected_matrix, file.path(data_dir, "06_DMSO_Corrected_Matrix_After_Imputation.xlsx"), rowNames = TRUE)
```

# STEP 11: ComBat Correction

```{r ComBat}
corrected_matrix <- ComBat(dat = dmso_corrected_matrix, batch = sample_info$Batch, mod = NULL, par.prior = TRUE)
# write.xlsx(corrected_matrix, file.path(data_dir, "07_ComBat_Corrected_Matrix.xlsx"), rowNames = TRUE)
```

# STEP 12: Visualization Analysis Before/After Correction (PCA)

```{r DMSO&Pool Groop}
# PCA on Corrected Matrix (DMSO & Pool Group)
pca_res_corrected <- prcomp(t(corrected_matrix), scale. = TRUE)
pca_data_corrected <- as.data.frame(pca_res_corrected$x) %>%
  rownames_to_column("FullSampleName") %>%
  left_join(sample_info, by = "FullSampleName")

percent_variance_corrected <- round(100 * pca_res_corrected$sdev^2 / sum(pca_res_corrected$sdev^2), 1)
pca_data_corrected <- pca_data_corrected %>%
  mutate(Group = case_when(
    str_detect(tolower(SampleType), "dmso") ~ "DMSO",
    str_detect(tolower(SampleType), "pool") ~ "Pool",
    TRUE                                     ~ "Drug Treatment"))

p_pca_groups <- ggplot(pca_data_corrected, aes(x = PC1, y = PC2)) +
  geom_point(
    data = pca_data_corrected %>% filter(Group == "Drug Treatment"),
    color = "#e5e5e5",
    size  = 2,
    alpha = 0.6
  ) +
  geom_point(
    data = pca_data_corrected %>% filter(Group %in% c("DMSO", "Pool")),
    aes(color = Group),
    size = 3
  ) +
  stat_ellipse(
    data = pca_data_corrected %>% filter(Group %in% c("DMSO", "Pool")),
    aes(color = Group),
    type  = "norm", 
    level = 0.95, 
    size  = 1
  ) +
  scale_color_manual(values = c("DMSO" = "#CC79A7", "Pool" = "#009E73")) +
  labs(
    title    = "PCA Plot (After Batch Correction)",
    subtitle = "95% confidence ellipses for DMSO and Pool samples",
    x = paste0("PC1 (", percent_variance_corrected[1], "% variance)"),
    y = paste0("PC2 (", percent_variance_corrected[2], "% variance)"),
    color = "Group"
  ) +
  theme_bw() +
  theme(
    legend.position = "right"
  ) + theme(aspect.ratio = 1)

print(p_pca_groups)

# PCA Comparison Before/After ComBat
plot_pca_custom <- function(mat, info, title) {
  pca <- prcomp(t(mat), scale. = TRUE)
  pct <- round(100 * pca$sdev^2 / sum(pca$sdev^2), 1)
  df <- as.data.frame(pca$x) %>%
    rownames_to_column("FullSampleName") %>%
    left_join(info, by = "FullSampleName")
  
  batches <- sort(unique(df$Batch))
  sorted <- batches[order(as.numeric(str_extract(batches, "\\d+")))]
  df$Batch <- factor(df$Batch, levels = sorted)
  
  ggplot(df, aes(x = PC1, y = PC2, color = Batch)) +
    geom_point(size = 2, alpha = 1) +
    labs(title = title, x = paste0("PC1 (", pct[1], "%)"), y = paste0("PC2 (", pct[2], "%)"))+
    theme_bw()
}

p_before <- plot_pca_custom(dmso_corrected_matrix, sample_info, "Before ComBat") + theme(legend.position = "bottom")
p_after <- plot_pca_custom(corrected_matrix, sample_info, "After ComBat") + theme(legend.position = "bottom")

combined_pca <- p_before + p_after + plot_layout(guides = "collect") + theme(aspect.ratio = 1)
print(combined_pca)
```
# optional

```{r}
plot_pca_custom <- function(mat, info, title) {
  info_nopool <- info %>%
    filter(!str_detect(tolower(SampleType), "pool"))
  keep_samples <- intersect(colnames(mat), info_nopool$FullSampleName)
  mat_sub  <- mat[, keep_samples, drop = FALSE]
  info_sub <- info_nopool %>% filter(FullSampleName %in% keep_samples)
  pca <- prcomp(t(mat_sub), scale. = TRUE)
  pct <- round(100 * pca$sdev^2 / sum(pca$sdev^2), 1)
  
  df <- as.data.frame(pca$x) %>%
    rownames_to_column("FullSampleName") %>%
    left_join(info_sub, by = "FullSampleName")
  batches <- sort(unique(df$Batch))
  sorted  <- batches[order(as.numeric(str_extract(batches, "\\d+")))]
  df$Batch <- factor(df$Batch, levels = sorted)
  
  ggplot(df, aes(x = PC1, y = PC2, color = Batch)) +
    geom_point(size = 2, alpha = 1) +
    labs(
      title = title,
      x = paste0("PC1 (", pct[1], "%)"),
      y = paste0("PC2 (", pct[2], "%)")
    ) +
    theme_bw()
}

p_before <- plot_pca_custom(dmso_corrected_matrix, sample_info, "Before ComBat") +
  theme(legend.position = "bottom")

p_after  <- plot_pca_custom(corrected_matrix,      sample_info, "After ComBat") +
  theme(legend.position = "bottom")

combined_pca <- p_before + p_after + plot_layout(guides = "collect")+theme(aspect.ratio = 1)

print(combined_pca)
#ggsave(file.path(fig_dir, "05_combinedPCA after batch correction_nopool.pdf"), combined_pca, width = 14, height = 7)
```

# STEP 13: Quality Control: Correlation Assessment (Pool / Technical / Biological Replicates)

```{r}
full_sample_cor_matrix <- cor(corrected_matrix, method = "spearman")

full_metadata <- tibble(
  FullName = colnames(corrected_matrix)
) %>%
  left_join(sample_info, by = c("FullName" = "FullSampleName")) %>%
  mutate(
    BaseDrugName = str_remove(str_extract(FullName, "(?<=_).*$"), "R$"),
    IsTechRep = str_ends(FullName, "R"),
    Category = case_when(
      str_detect(tolower(SampleType), "pool") ~ "Pool",
      str_detect(tolower(SampleType), "dmso") ~ "DMSO",
      TRUE                                     ~ "Drug"
    )
  )

## 1. Pool Sample Correlation --------------------------------------------------------

pool_samples   <- full_metadata %>%
  filter(Category == "Pool") %>%
  pull(FullName)

pool_cor_subset <- full_sample_cor_matrix[pool_samples, pool_samples]
pool_cor_values <- pool_cor_subset[upper.tri(pool_cor_subset)]
median_pool_cor <- median(pool_cor_values)

cat("\n--- 1. Pool Sample QC (Platform Stability) ---\n")
cat(paste("Found", length(pool_samples), "Pool samples.\n"))
cat(paste("Median Correlation:", round(median_pool_cor, 4), "\n"))
if (median_pool_cor > 0.95) cat("Assessment: Excellent, platform stability is very high.\n")

# Sort Pool samples by Batch number
pool_order <- full_metadata %>%
  filter(Category == "Pool") %>%
  mutate(Batch_num = as.numeric(str_extract(Batch, "\\d+"))) %>%
  arrange(Batch_num) %>%
  pull(FullName)

pool_cor_sorted <- pool_cor_subset[pool_order, pool_order]

# Corrplot visualization
pdf(file.path(fig_dir, "08_Pool_Correlation.pdf"), width = 20, height = 10)
corrplot(
  pool_cor_sorted,
  method  = "circle",
  type    = "upper",
  order   = "original",
  col     = colorRampPalette(c("#3462ae", "white", "#9e539d"))(100),
  tl.col  = "black",
  tl.srt  = 90,
  tl.cex  = 0.8,
  diag    = FALSE,
  addgrid.col   = "grey90",
  addCoef.col   = "black",
  number.cex    = 0.4,
  number.digits = 2,
  title         = "Correlation of Pool Samples",
  mar           = c(0, 0, 2, 0)
)
dev.off()

## 2. Technical Replicate Correlation --------------------------------------------------------

base_samples     <- full_metadata %>% filter(Category == "Drug", !IsTechRep)
tech_rep_samples <- full_metadata %>% filter(Category == "Drug",  IsTechRep)

tech_replicates_pairs <- inner_join(
  base_samples, tech_rep_samples,
  by = c("Batch", "BaseDrugName"),
  suffix = c("_base", "_rep")
)

tech_replicates_cor <- tech_replicates_pairs %>%
  rowwise() %>%
  mutate(
    Correlation = full_sample_cor_matrix[FullName_base, FullName_rep]
  ) %>%
  ungroup() %>%
  na.omit()

median_tech_rep_cor <- median(tech_replicates_cor$Correlation)

cat("\n--- 2. Technical Replicate QC (Processing Stability) ---\n")
cat(paste("Found", nrow(tech_replicates_cor), "technical replicate pairs.\n"))
cat(paste("Median Correlation:", round(median_tech_rep_cor, 4), "\n"))
if (median_tech_rep_cor > 0.90) cat("Assessment: Excellent, processing workflow highly reproducible.\n")

## 3. Biological Replicate Correlation ------------------------------------------------------

drug_samples_meta <- full_metadata %>% filter(Category == "Drug")

bio_replicates_pairs <- drug_samples_meta %>%
  inner_join(drug_samples_meta, by = "BaseDrugName", suffix = c("1", "2")) %>%
  filter(FullName1 < FullName2) %>%  
  filter(Batch1   != Batch2)  

bio_replicates_cor <- bio_replicates_pairs %>%
  rowwise() %>%
  mutate(
    Correlation = full_sample_cor_matrix[FullName1, FullName2]
  ) %>%
  ungroup() %>%
  na.omit()

median_bio_rep_cor <- median(bio_replicates_cor$Correlation)

cat("\n--- 3. Biological Replicate QC (Overall Experimental Stability) ---\n")
cat(paste("Found", nrow(bio_replicates_cor), "biological replicate pairs.\n"))
cat(paste("Median Correlation:", round(median_bio_rep_cor, 4), "\n"))
if (median_bio_rep_cor > 0.75) cat("Assessment: Excellent, high consistency across batches.\n")

## 4. Comprehensive QC Violin Plot ------------------------------------------------------

qc_data_pool <- tibble(Correlation = pool_cor_values, Type = "Pool vs. Pool")
qc_data_tech <- tibble(Correlation = tech_replicates_cor$Correlation, Type = "Technical Replicates")
qc_data_bio  <- tibble(Correlation = bio_replicates_cor$Correlation, Type = "Biological Replicates")

combined_qc_data <- bind_rows(qc_data_pool, qc_data_tech, qc_data_bio) %>%
  mutate(Type = factor(Type, levels = c("Pool vs. Pool", "Technical Replicates", "Biological Replicates")))

p_cor_violin <- ggplot(combined_qc_data, aes(x = Type, y = Correlation, fill = Type)) +
  geom_violin(trim = FALSE, alpha = 0.8) +
  geom_boxplot(width = 0.15, fill = "white", outlier.size = 1) +
  labs(
    title    = "Comprehensive Quality Control via Sample Correlation",
    subtitle = "Distribution of Spearman correlations for different replicate types",
    x        = "Comparison Type",
    y        = "Spearman Correlation"
  ) +
  scale_fill_manual(values = c(
    "Pool vs. Pool"         = "#009E73",
    "Technical Replicates"  = "#F0E442",
    "Biological Replicates" = "#0072B2"
  )) +
  stat_summary(fun = median, geom = "text", aes(label = sprintf("%.3f", ..y..)), vjust = -0.5, size = 4.5, color = "black") +
  theme_bw(base_size = 16) +
  theme(axis.text.x = element_text(angle = 10, hjust = 1), legend.position = "none") +
  coord_cartesian(ylim = c(floor(min(combined_qc_data$Correlation) * 10) / 10, 1.0))

print(p_cor_violin)

## 5. Pool Correlation Heatmap ------------------------------------------------------

cor_long <- melt(pool_cor_sorted, varnames = c("Sample1", "Sample2"), value.name = "Correlation")
cor_long$Sample1 <- factor(cor_long$Sample1, levels = pool_order)
cor_long$Sample2 <- factor(cor_long$Sample2, levels = pool_order)

p_pool_heat <- ggplot(cor_long, aes(x = Sample1, y = Sample2, fill = Correlation)) +
  geom_tile(color = "grey90") +
  scale_fill_gradient2(low = "#3462ae", mid = "white", high = "#9e539d", midpoint = 0.8, limits = c(0.6, 1), name = "Spearman r") +
  coord_fixed() +
  labs(title = sprintf("Correlation of Pool Samples (median = %.3f)", median_pool_cor), x = NULL, y = NULL) +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6), axis.text.y = element_text(size = 6), panel.grid = element_blank()) + theme(aspect.ratio = 1)

print(p_pool_heat)
```

# STEP 14: QC Based on CV (Pool / Technical / Biological Replicates)

```{r}
linear_matrix <- 2^corrected_matrix
cat("Data converted back to linear scale for CV calculation.\n")

calculate_cv <- function(x) {
  mean_x <- mean(x, na.rm = TRUE)
  if (mean_x == 0) return(NA)
  (sd(x, na.rm = TRUE) / mean_x) * 100
}

## 1. Pool Sample CV --------------------------------------------------------

pool_samples      <- full_metadata %>% filter(Category == "Pool") %>% pull(FullName)
pool_linear_subset <- linear_matrix[, pool_samples]
pool_cv_values <- apply(pool_linear_subset, 1, calculate_cv) %>% na.omit()
median_pool_cv <- median(pool_cv_values)

cat("\n--- 1. Pool Sample CV (Platform Precision) ---\n")
cat(paste("Median CV (%):", round(median_pool_cv, 2), "%\n"))

## 2. Technical Replicate CV ---------------------------------------------------------

tech_rep_cv_list <- list()
for (i in 1:nrow(tech_replicates_pairs)) {
  pair <- tech_replicates_pairs[i, ]
  sample_names     <- c(pair$FullName_base, pair$FullName_rep)
  pair_linear_data <- linear_matrix[, sample_names]
  tech_rep_cv_list[[i]] <- apply(pair_linear_data, 1, calculate_cv)
}
tech_rep_cv_values <- unlist(tech_rep_cv_list) %>% na.omit()
median_tech_rep_cv <- median(tech_rep_cv_values)

cat("\n--- 2. Technical Replicate CV (Processing Precision) ---\n")
cat(paste("Median CV (%):", round(median_tech_rep_cv, 2), "%\n"))

## 3. Biological Replicate CV -------------------------------------------------------

replicated_drugs <- drug_samples_meta %>% group_by(BaseDrugName) %>% filter(n() > 1) %>% distinct(BaseDrugName) %>% pull(BaseDrugName)

bio_rep_cv_list <- list()
for (drug in replicated_drugs) {
  sample_names <- drug_samples_meta %>% filter(BaseDrugName == drug) %>% pull(FullName)
  group_linear_data <- linear_matrix[, sample_names]
  bio_rep_cv_list[[drug]] <- apply(group_linear_data, 1, calculate_cv)
}
bio_rep_cv_values <- unlist(bio_rep_cv_list) %>% na.omit()
median_bio_rep_cv <- median(bio_rep_cv_values)

cat("\n--- 3. Biological Replicate CV (Overall Experimental Precision) ---\n")
cat(paste("Median CV (%):", round(median_bio_rep_cv, 2), "%\n"))

## 4. Comprehensive CV Violin Plot ------------------------------------------------------

qc_cv_pool <- tibble(CV = pool_cv_values, Type = "Pool Samples")
qc_cv_tech <- tibble(CV = tech_rep_cv_values, Type = "Technical Replicates")
qc_cv_bio  <- tibble(CV = bio_rep_cv_values, Type = "Biological Replicates")

combined_cv_data <- bind_rows(qc_cv_pool, qc_cv_tech, qc_cv_bio) %>%
  mutate(Type = factor(Type, levels = c("Pool Samples", "Technical Replicates", "Biological Replicates")))

p_cv_violin <- ggplot(combined_cv_data, aes(x = Type, y = CV, fill = Type)) +
  geom_violin(trim = FALSE, alpha = 0.8) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  labs(
    title    = "Comprehensive Quality Control via Coefficient of Variation (CV)",
    subtitle = "CV calculated on linear-scale data. Lower CV indicates higher precision.",
    x        = "Comparison Type",
    y        = "Coefficient of Variation (%)"
  ) +
  scale_fill_manual(values = c("Pool Samples" = "#009E73", "Technical Replicates" = "#F0E442", "Biological Replicates" = "#0072B2")) +
  theme_bw(base_size = 14) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1), legend.position = "none") +
  coord_cartesian(ylim = c(0, 150)) +
  stat_summary(fun = median, geom = "text", aes(label = paste0(round(..y.., 1), "%")), vjust = -0.5, size = 4)

print(p_cv_violin)

## Technical vs Biological Replicates Only -----------------------------------

qc_cv_tech2 <- tibble(CV = tech_rep_cv_values, Type = "Technical Replicates")
qc_cv_bio2  <- tibble(CV = bio_rep_cv_values,  Type = "Biological Replicates")

combined_cv_data2 <- bind_rows(qc_cv_tech2, qc_cv_bio2) %>%
  mutate(Type = factor(Type, levels = c("Technical Replicates", "Biological Replicates")))

p_cv_tech_bio <- ggplot(combined_cv_data2, aes(x = Type, y = CV, fill = Type)) +
  geom_violin(trim = FALSE, alpha = 0.8) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  labs(title = "CV of Technical vs Biological Replicates", y = "Coefficient of Variation (%)") +
  scale_fill_manual(values = c("Technical Replicates" = "#3462ae", "Biological Replicates" = "#9e539d")) +
  theme_bw(base_size = 14) +
  theme(axis.text.x = element_text(angle = 15, hjust = 1), legend.position = "none") +
  coord_cartesian(ylim = c(0, 110)) +
  stat_summary(fun = median, geom = "text", aes(label = paste0(round(..y.., 1), "%")), vjust = -0.5, size = 4) + theme(aspect.ratio = 1)

print(p_cv_tech_bio)
#write.xlsx(combined_cv_data2, file.path(data_dir, "Plot04_combined_cv_data2.xlsx"))
#ggsave(file.path(fig_dir, "04_cv of technical and Biological replicates.pdf"), p_cv_tech_bio, width = 7, height = 7)
```

# STEP 15: Log2FC Calculation (Global DMSO Median)

```{r}
all_dmso_samples <- sample_info %>% filter(str_detect(tolower(SampleType), "dmso")) %>% pull(FullSampleName)
all_drug_samples_info <- sample_info %>% filter(!str_detect(tolower(SampleType), "pool|dmso"))

corrected_dmso_matrix <- corrected_matrix[, all_dmso_samples]
global_dmso_profile_corrected <- apply(corrected_dmso_matrix, 1, median, na.rm = TRUE)
all_drug_sample_names <- all_drug_samples_info %>% pull(FullSampleName)
corrected_drugs_matrix <- corrected_matrix[, all_drug_sample_names]

l2fc_matrix_combat_global_dmso <- sweep(corrected_drugs_matrix, 1, global_dmso_profile_corrected, FUN = "-")
short_drug_names <- paste(all_drug_samples_info$Batch, all_drug_samples_info$SampleType, sep="_")
colnames(l2fc_matrix_combat_global_dmso) <- short_drug_names

# write.xlsx(l2fc_matrix_combat_global_dmso, file.path(data_dir, "09_Log2FC_Global_DMSO_Median.xlsx"), rowNames=TRUE)
```

# STEP 16: Log2FC Replicate Correlation (Technical & Biological)

```{r}
final_l2fc_matrix <- l2fc_matrix_combat_global_dmso

# Correlation matrix of Log2FC
l2fc_cor_matrix <- cor(final_l2fc_matrix, method = "spearman", use = "pairwise.complete.obs")

# Technical Replicates on Log2FC
tech_replicates_cor_l2fc <- tech_replicates_pairs %>%
  rowwise() %>% mutate(Correlation_l2fc = l2fc_cor_matrix[FullName_base, FullName_rep]) %>% ungroup() %>% na.omit()
median_tech_rep_cor_l2fc <- median(tech_replicates_cor_l2fc$Correlation_l2fc)

cat("\n--- Technical Replicate Correlation on Log2FC Matrix ---\n")
cat("Pairs:", nrow(tech_replicates_cor_l2fc), "\n")
cat("Median Correlation:", round(median_tech_rep_cor_l2fc, 4), "\n")

# Biological Replicates on Log2FC
bio_replicates_cor_l2fc <- bio_replicates_pairs %>%
  rowwise() %>% mutate(Correlation_l2fc = l2fc_cor_matrix[FullName1, FullName2]) %>% ungroup() %>% na.omit()
median_bio_rep_cor_l2fc <- median(bio_replicates_cor_l2fc$Correlation_l2fc)

cat("\n--- Biological Replicate Correlation on Log2FC Matrix ---\n")
cat("Pairs:", nrow(bio_replicates_cor_l2fc), "\n")
cat("Median Correlation:", round(median_bio_rep_cor_l2fc, 4), "\n")

# Violin Plot
qc_rep_corr <- bind_rows(
  tech_replicates_cor_l2fc %>% transmute(Type = "Technical Replicates", Correlation = Correlation_l2fc),
  bio_replicates_cor_l2fc %>% transmute(Type = "Biological Replicates", Correlation = Correlation_l2fc)
) %>% mutate(Type = factor(Type, levels = c("Technical Replicates", "Biological Replicates")))

p_l2fc_qc <- ggplot(qc_rep_corr, aes(x = Type, y = Correlation, fill = Type)) +
  geom_violin(trim = FALSE, alpha = 0.8, scale = "width") +
  geom_boxplot(width = 0.12, fill = "white", outlier.shape = NA, alpha = 0.9) +
  scale_fill_manual(values = c("Technical Replicates" = "#3462ae", "Biological Replicates" = "#9e539d")) +
  stat_summary(fun = median, geom = "text", aes(label = sprintf("%.3f", after_stat(y))), vjust = -0.8, size = 5, fontface = "bold") +
  labs(title = "Replicate Correlation in Log2FC Matrix", subtitle = "QC assessment for technical and biological replicates", x = NULL, y = "Spearman Correlation") +
  theme_bw() + theme(axis.text.x = element_text(size = 12), legend.position = "none", plot.title = element_text(face = "bold", size = 14)) +
  coord_cartesian(ylim = c(0, 1.1)) + theme(aspect.ratio = 1)

print(p_l2fc_qc)
```

# STEP 17: Absolute Log2FC Difference Analysis Between Biological Replicates

```{r}
# Metadata for L2FC matrix
l2fc_metadata <- tibble(FullName = colnames(final_l2fc_matrix)) %>%
  mutate(Batch = str_extract(FullName, "^[^_]+"), DrugPart = str_extract(FullName, "(?<=_).*$"),
         BaseDrugName = str_remove(DrugPart, "R$"), IsTechRep = str_ends(DrugPart, "R"))

# Find biological replicate pairs
bio_rep_pairs <- l2fc_metadata %>%
  inner_join(l2fc_metadata, by = "BaseDrugName", suffix = c("_1", "_2")) %>%
  filter(FullName_1 < FullName_2, Batch_1 != Batch_2)

cat("Found", nrow(bio_rep_pairs), "biological replicate pairs\n")

# Calculate differences
all_differences <- list()
for(i in 1:nrow(bio_rep_pairs)) {
  pair <- bio_rep_pairs[i, ]
  l2fc_rep1 <- final_l2fc_matrix[, pair$FullName_1]
  l2fc_rep2 <- final_l2fc_matrix[, pair$FullName_2]
  abs_diff <- abs(l2fc_rep1 - l2fc_rep2)
  all_differences[[i]] <- tibble(Protein = rownames(final_l2fc_matrix), Drug = pair$BaseDrugName, Rep1 = pair$FullName_1, Rep2 = pair$FullName_2, L2FC_Rep1 = l2fc_rep1, L2FC_Rep2 = l2fc_rep2, Abs_Difference = abs_diff)
}

diff_data <- bind_rows(all_differences) %>% filter(!is.na(Abs_Difference))

# Statistics
thresholds <- c(0.25, 0.5, 0.75, 1.0)
for(thresh in thresholds) {
  pct <- mean(diff_data$Abs_Difference > thresh) * 100
  cat(sprintf("  > %.2f: %.2f%% (%d pairs)\n", thresh, pct, sum(diff_data$Abs_Difference > thresh)))
}

# Density Plot
p_density <- ggplot(diff_data, aes(x = Abs_Difference)) +
  geom_density(fill = "#3462ae", alpha = 0.6) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "red", size = 1) +
  geom_vline(xintercept = median(diff_data$Abs_Difference), linetype = "solid", color = "#9e539d", size = 1) +
  annotate("text", x = median(diff_data$Abs_Difference) + 0.02, y = Inf, vjust = 2, hjust = 0, label = paste0("Median = ", round(median(diff_data$Abs_Difference), 3)), color = "#9e539d", size = 4) +
  labs(title = "Density of Log2FC Differences Between Biological Replicates", subtitle = paste0(sprintf("%.1f%%", mean(diff_data$Abs_Difference > 0.5) * 100), " of measurements differ by more than 0.5"), x = "Absolute Difference in Log2FC", y = "Density") +
  theme_bw() + theme(plot.title = element_text(face = "bold"), panel.grid.minor = element_blank()) + coord_cartesian(xlim = c(0, 2)) + theme(aspect.ratio = 1)

print(p_density)

# Summary Report
cat("\n========== Biological Replicate Consistency Assessment Report ==========\n")
cat(sprintf("Total protein-compound measurement pairs: %s\n", format(nrow(diff_data), big.mark = ",")))
cat(sprintf("Median Log2FC Difference: %.4f\n", median(diff_data$Abs_Difference)))
cat(sprintf("Proportion Difference <= 0.5: %.2f%%\n", mean(diff_data$Abs_Difference <= 0.5) * 100))
cat(sprintf("Proportion Difference > 0.5: %.2f%%\n", mean(diff_data$Abs_Difference > 0.5) * 100))

summary_stats <- tibble(Metric = c("Total protein-compound pairs", "Median absolute difference", "Mean absolute difference", "SD of difference", "% within 0.25", "% within 0.5", "% within 0.75", "% within 1.0"), Value = c(nrow(diff_data), round(median(diff_data$Abs_Difference), 4), round(mean(diff_data$Abs_Difference), 4), round(sd(diff_data$Abs_Difference), 4), round(mean(diff_data$Abs_Difference <= 0.25) * 100, 2), round(mean(diff_data$Abs_Difference <= 0.5) * 100, 2), round(mean(diff_data$Abs_Difference <= 0.75) * 100, 2), round(mean(diff_data$Abs_Difference <= 1.0) * 100, 2)))

# write.xlsx(summary_stats, file.path(data_dir, "10_BioRep_L2FC_Difference_Summary_Stats.xlsx"))
#write.xlsx(diff_data, file.path(data_dir, "Plot03_diff_data.xlsx"))
#ggsave(file.path(fig_dir, "03_density of log2fc between BR.pdf"), p_density, width = 7, height = 7)
```

# STEP 18: Merge Replicates (Two-step averaging) to Generate Final Matrix

```{r}
full_metadata <- tibble(FullName = colnames(corrected_matrix)) %>% 
  left_join(sample_info, by = c("FullName" = "FullSampleName")) %>% 
  mutate(BaseDrugName = str_remove(str_extract(FullName, "(?<=_).*$"), "R$"),
         IsTechRep = str_ends(FullName, "R"),
         Category = case_when(str_detect(tolower(SampleType), "pool") ~ "Pool", str_detect(tolower(SampleType), "dmso") ~ "DMSO", TRUE ~ "Drug"),
         GroupToMerge = case_when(Category == "Drug" ~ BaseDrugName, Category == "DMSO" ~ "DMSO_Control", Category == "Pool" ~ "Pool_Control"))

unique_groups <- unique(full_metadata$GroupToMerge)
cat(paste("Found", length(unique_groups), "independent groups for merging.\n"))

merged_vectors_list <- list()
for (group_name in unique_groups) {
  replicate_cols <- full_metadata %>% filter(GroupToMerge == group_name) %>% pull(FullName)
  replicates_subset_matrix <- corrected_matrix[, replicate_cols, drop = FALSE]
  merged_vector <- rowMeans(replicates_subset_matrix, na.rm = TRUE)
  merged_vectors_list[[group_name]] <- merged_vector
}

merged_matrix <- do.call(cbind, merged_vectors_list)

cat("\n--- Replicate Merging Completed ---\n")
print(paste("Original Matrix Dim:", paste(dim(corrected_matrix), collapse = " x ")))
print(paste("Merged Matrix Dim:", paste(dim(merged_matrix), collapse = " x ")))

# write.xlsx(merged_matrix, file.path(data_dir, "11_Merged_Matrix.xlsx"), rowNames=TRUE)
```

# STEP 19: Calculate Final Robust L2FC Matrix

```{r}
dmso_control_vector <- merged_matrix[, "DMSO_Control"]
final_l2fc_matrix <- sweep(merged_matrix, 1, dmso_control_vector, FUN = "-")
drug_columns <- setdiff(colnames(final_l2fc_matrix), c("DMSO_Control", "Pool_Control"))
final_l2fc_matrix_drugs_only <- final_l2fc_matrix[, drug_columns]

cat("--- Final L2FC Matrix Calculation Completed ---\n")
print(paste("Final L2FC Matrix (Drugs Only) Dim:", paste(dim(final_l2fc_matrix_drugs_only), collapse = " x ")))
print("Remaining Column Names Sample:")
print(head(colnames(final_l2fc_matrix_drugs_only)))

write.xlsx(final_l2fc_matrix_drugs_only, file.path(data_dir, "12_Final_L2FC_Matrix_Drugs_Only_367.xlsx"), rowNames=TRUE)
cat("\n>>> Data processing Successfully <<<\n")
```



# Figure S1A_Sankey plot
```{r}
library(dplyr)
library(ggplot2)
library(ggalluvial)
library(tidyr)
library(openxlsx)

# Load data
df <- read.xlsx(file.path(input_dir, "Sankey_Plot.xlsx")) 

# Step 1: Preprocessing ----
data_sankey <- df %>%
  # Convert to long format first
  separate_rows(Pathways, sep = ";") %>%
  to_lodes_form(
    axes = c("Class1", "Drug", "Class2", "Class3", "Pathways"),
    key = "layer",
    id = "id"
  ) %>%
  # Regenerate Class1 column
  mutate(
    Class1 = if_else(layer == "Class1", stratum, NA_character_)
  ) %>%
  # Fill Class1 values by id
  group_by(id) %>%
  fill(Class1, .direction = "downup") %>%  # Bi-directional fill to ensure coverage
  ungroup() %>%
  mutate(Class1 = factor(Class1))  # Convert to factor


# Step 2: Calculate proportions correctly ----
layer_totals <- data_sankey %>%
  group_by(layer) %>%
  summarise(total = n())

data_sankey <- data_sankey %>%
  group_by(layer, stratum) %>%
  mutate(percent = n() / layer_totals$total[layer_totals$layer == cur_group()$layer] * 100) %>%
  ungroup()

# Step 3: Highlight target category ----
highlight_class <- "RTK"
highlight_layer <- "Class2"

# Get full flow paths for the highlighted IDs
highlight_ids <- data_sankey %>% 
  filter(layer == highlight_layer & stratum == highlight_class) %>% 
  pull(id) %>% 
  unique()

# Step 4: Plot Optimization ----
n_categories <- data_sankey %>% 
  pull(Class1) %>% 
  na.omit() %>%
  unique() %>% 
  length()

# Generate purple gradient
purple_gradient <- colorRampPalette(
  c("#9b9ac1",  "#c2c8d8", "#b7a6c7", "#cabad7")  # Light to dark purple palette
)(n_categories)

# Main Plot Code
p <- ggplot(data_sankey, aes(x = layer, stratum = stratum, alluvium = id)) +
  geom_flow(
    aes(fill = Class1),
    width = 0.25,
    alpha = 0.15,
    color = "grey80",
    curve_type = "cubic"
  ) +
  geom_flow(
    data = data_sankey %>% filter(id %in% highlight_ids),
    aes(fill = Class1),
    width = 0.25,
    alpha = 0.6,
    color = "red",
    curve_type = "cubic"
  ) +
  geom_stratum(
    aes(fill = Class1),
    width = 0.3,
    alpha = 0.6,
    color = "NA"
  ) +
  geom_stratum(
    data = data_sankey %>% filter(stratum == highlight_class),
    aes(fill = Class1),
    width = 0.3,
    alpha = 0.9,
    color = "red",
    linewidth = 0.1
  ) +
  geom_text(
    stat = "stratum",
    aes(label = ifelse(percent > 1, 
                       paste0(stratum, "\n", round(percent,1), "%"), 
                       NA)),
    size = 2.8,
    color = "black",
    vjust = -0.6,
    lineheight = 0.8
  ) +
  # Purple gradient scale
  scale_fill_manual(
    values = purple_gradient,
    na.value = "#D0D2D4",
    guide = guide_legend(
      title = "Class 1",
      override.aes = list(alpha = 0.8, color = "gray30")
    )
  )  +
  theme_void() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold", size = 13)
  )
p

#### Proportion Calculation and Export ----
library(writexl)  

# Generate hierarchical percentage summary table
percentage_summary <- data_sankey %>%
  # Group by layer and stratum
  group_by(layer, stratum) %>%
  # Extract unique values to avoid duplicate counting
  summarise(
    percent = first(percent),  # Percentage pre-calculated
    Class1 = first(Class1),    # Associated Class1
    .groups = "drop"
  ) %>%
  # Sort by layer and percentage
  arrange(layer, desc(percent)) %>%
  # Rename columns to English
  rename(
    "Category_Layer" = layer,
    "Specific_Category" = stratum,
    "Percentage_Percent" = percent,
    "Primary_Class" = Class1
  )

library(readr)
# Export to CSV (Statistics of proportions for each category)
write_excel_csv(
  percentage_summary, 
  file = file.path(data_dir, "Drug_Classification_Proportions.csv"),
  na = ""  # Replace NA with empty string
)
```

# Figure 1A
```{r}
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(cluster)
library(rcdk)
library(factoextra)
library(itertools)
library(pvclust)
library(dendextend)
library(circlize)
library(here)

# https://jcheminf.biomedcentral.com/track/pdf/10.1186/s13321-019-0405-0.pdf

#### 1. Import Files ####
# Using GB18030 encoding to prevent errors with special characters (like ±)
smiles_367 <- read.delim(file.path(input_dir, "367drugs.smiles.txt"), header = FALSE, fileEncoding = "GB18030")
colnames(smiles_367) <- c("smiles", "drug", "drug_first")
rownames(smiles_367) <- smiles_367$drug

mols <- parse.smiles(smiles_367$smiles)
names(mols) <- rownames(smiles_367)

## Get fingerprints
# 'extended' type as suggested by Voicu et al.
fp <- lapply(mols, get.fingerprint, type = 'extended') 
fp.sim <- fingerprint::fp.sim.matrix(fp, method = 'tanimoto')
fp.dist <- 1 - fp.sim

rownames(fp.dist) <- rownames(smiles_367)
colnames(fp.dist) <- rownames(smiles_367)

#### 2. Evaluate Optimal Clustering Values ####

# Clustering Method 1: Gap Statistic
# Adjusting k.max; a higher number of clusters often appears slightly better
fviz_nbclust(fp.dist, hcut, diss = T, method = "gap_stat", k.max = 300)

# ============================================================
# Extract optimal k from fviz_nbclust results across different k.max
kmax_values <- c(20, 50, 80, 100, 150, 200, 250)
optimal_k <- c()

cat("Calculating optimal k for different k.max values...\n")

for (kmax in kmax_values) {
  result <- fviz_nbclust(fp.dist, hcut, diss = TRUE, 
                         method = "gap_stat", k.max = kmax)
  
  # Extract data from ggplot object
  plot_data <- result$data
  best_k <- which.max(plot_data$gap)
  optimal_k <- c(optimal_k, best_k)
  cat("  k.max =", kmax, "→ optimal k =", best_k, "\n")
}

# Bind results
result_df <- data.frame(k_max = kmax_values, optimal_k = optimal_k)

p <- ggplot(result_df, aes(x = k_max, y = optimal_k)) +
  geom_line(color = "red", linewidth = 1.5) +
  geom_point(color = "red", size = 4) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  annotate("text", x = 120, y = 80,
           label = "Optimal k ≈ k.max\n→ Does not converge\n→ No natural clusters",
           size = 4.5, color = "red") +
  labs(title = "Gap Statistic: Optimal k vs k.max",
       subtitle = "Optimal k increases linearly with k.max → No convergence",
       x = "k.max (maximum clusters tested)",
       y = "Optimal k (suggested by Gap statistic)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))

print(p)


# Clustering Method 2: Silhouette
k_values <- 2:366

results <- data.frame(k = k_values, silhouette = NA)
for (i in seq_along(k_values)) {
  k <- k_values[i]
  hc <- hclust(as.dist(fp.dist), method = "complete")
  clusters <- cutree(hc, k = k)
  sil <- silhouette(clusters, as.dist(fp.dist))
  results$silhouette[i] <- mean(sil[, 3])
  
  if (k %% 50 == 0) cat("  k =", k, "completed\n")
}

# Identify best k
best_k <- results$k[which.max(results$silhouette)]
best_sil <- max(results$silhouette)

# Silhouette plot
q <- ggplot(results, aes(x = k, y = silhouette)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(data = results[results$k == best_k, ], 
             aes(x = k, y = silhouette),
             color = "red", size = 4) +
  geom_vline(xintercept = best_k, linetype = "dashed", color = "red") +
  annotate("text", x = best_k + 10, y = best_sil, 
           label = paste0("Optimal k = ", best_k, "\n(", 
                          round(367/best_k, 1), " drugs/cluster)"),
           hjust = 0, color = "red", size = 4) +
  annotate("text", x = 150, y = 0.15,
           label = "Optimal k = 220 (≈1.7 drugs/cluster),\nindicating that no small number of well-separated structure-based clusters can be identified.",
           size = 4, color = "gray40") +
  labs(title = "Chemical Structure-based Clustering of 367 Anticancer Drugs",
       subtitle = "Silhouette analysis reveals lack of natural cluster structure",
       x = "Number of Clusters (k)",
       y = "Average Silhouette Width") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 11, color = "gray40"))

print(q)


#### 3. Evaluate Clustering Stability ####

# Evaluate branch stability via Bootstrap
drug_clust_boot <- pvclust(fp.dist, method.hclust = "complete",
                           method.dist = "euclidean", nboot = 100,
                           iseed = 2248)

k_220 <- cutree(drug_clust_boot, k = 220, order_clusters_as_data = FALSE)


#### 4. Plotting Preliminary Drug Clusters vs ATC Codes ####
# Drug information (ATC classes)
drug_meta <- read.csv(file.path(input_dir, "Drugs_sampleinfo_367.csv"), row.names = 1) 
drug_meta <- merge(drug_meta, smiles_367, by.x = "Drug_first",
                   by.y = "drug_first", all.x = T) %>%
  mutate(ATC_colours = case_when(str_detect(Level_1, 'A') ~ '#f4c40f',
                                 str_detect(Level_1, 'B') ~ '#fe9b00',
                                 str_detect(Level_1, 'C') ~ '#d8443c',
                                 str_detect(Level_1, 'D') ~ '#9b3441',
                                 str_detect(Level_1, 'G') ~ '#de597c',
                                 str_detect(Level_1, 'H') ~ '#F0A8B2',
                                 str_detect(Level_1, 'J') ~ '#aa7aa1',
                                 str_detect(Level_1, 'L') ~ '#9f5691',
                                 str_detect(Level_1, 'M') ~ '#633372',
                                 str_detect(Level_1, 'N') ~ '#1f6e9c',
                                 str_detect(Level_1, 'R') ~ '#2b9b81',
                                 str_detect(Level_1, 'S') ~ '#92c051',
                                 str_detect(Level_1, 'V') ~ 'black',
                                 str_detect(Level_1, 'P') ~ 'red',
                                 TRUE ~ 'grey'))

drug_meta <- drug_meta[match(names(k_220), drug_meta$Drug_first), ]
drug_meta$ATC_colours[is.na(drug_meta$ATC_colours)] <- "grey"

# Define Legend
library(dendextend)
library(circlize)

# Prepare mapping data
class <- c('A', 'B', 'C', 'D', 'G', 'H', 'J', 'L', 
           'M', 'N', 'R', 'S', 'P', 'V', "NA")
colours <- c('#f4c40f','#fe9b00','#d8443c','#9b3441','#de597c','#F0A8B2',
             '#aa7aa1','#9f5691','#633372','#1f6e9c','#2b9b81','#92c051',
             'red','black', 'grey')

dend <- as.dendrogram(drug_clust_boot)


dend_styled <- dend %>%
  set("leaves_pch", NA) %>%           # Set leaf shape to NA (hidden)
  set("leaves_cex", 0) %>%            # Set leaf size to 0
  set("labels_cex", 0) %>%            # Hide original labels (drawn manually)
  color_branches(k = 220) %>%
  pvclust_show_signif(
    drug_clust_boot,
    alpha         = 0.05,
    signif_type   = "au",
    lwd_signif    = 0.04,
    lwd_nonsignif = 0.01
  )

n_leaves <- nleaves(dend)
leaf_labels <- labels(dend)
leaf_colors <- drug_meta$ATC_colours[match(leaf_labels, drug_meta$Drug_first)]
leaf_colors[is.na(leaf_colors)] <- "grey"

# ============================================================
# Start Plotting
# ============================================================

pdf(file.path(fig_dir, "Figure1A_drugstructure_hclust_circlize_k220_manual.pdf"), width = 20, height = 17)

circos.clear()

circos.par(
  cell.padding = c(0, 0, 0, 0),
  track.margin = c(0, 0),
  start.degree = 90,
  gap.degree = 0,
  canvas.xlim = c(-1.2, 1.2),
  canvas.ylim = c(-1.2, 1.2)
)

circos.initialize(factors = "a", xlim = c(0, n_leaves))

# Track 1: Labels
circos.track(
  ylim = c(0, 1),
  track.height = 0.04,
  bg.border = NA,
  panel.fun = function(x, y) {
    for (i in 1:n_leaves) {
      x_pos <- i - 0.5
      circos.text(
        x = x_pos,
        y = 0.5,
        labels = leaf_labels[i],
        facing = "clockwise",
        niceFacing = TRUE,
        adj = c(0, 0.5),
        cex = 0.8
      )
    }
  }
)

# Track 2: Leaf points (The colored ring)
circos.track(
  ylim = c(0, 1),
  track.height = 0.03,
  bg.border = NA,
  panel.fun = function(x, y) {
    for (i in 1:n_leaves) {
      x_pos <- i - 0.5
      circos.points(
        x = x_pos,
        y = 0.5,
        pch = 19,
        cex = 1,
        col = leaf_colors[i]
      )
    }
  }
)

# Track 3: Spacing
circos.track(
  ylim = c(0, 1),
  track.height = 0.0005,
  bg.border = NA
)

# Track 4: Dendrogram (Leaves are hidden here)
circos.track(
  ylim = c(0, attr(dend_styled, "height")),
  track.height = 0.4,
  bg.border = NA,
  panel.fun = function(x, y) {
    circos.dendrogram(
      dend_styled,
      facing = "outside"
    )
  }
)

legend("topleft",
       legend = class,
       fill   = colours,
       cex    = 0.9,
       ncol   = 2,
       title  = "ATC Level-1 Class",
       bty    = "n")


dev.off()
circos.clear()

cat("PDF has been saved!\n")
```

